Using Occupancy Theory and Hierarchical Models to Estimate Otter Population in Southeastern Minnesota

### 2010-02-25 Paper.tex
added figure, edited false positive section

    \subsection{Descriptive statistics}
    In 2003, three different observers flew over the rivers. Assuming that no
    tracks are laid down between individuals' flights on the same day, all three
    people should have seen the same tracks. While there is some error in
    classifying which site the observer saw the actual track in, each observer
    should at least be seeing the same, or nearly the same, number of sites with
    tracks. However, we did not find that to be the case. If we look only at
    days that all three people observed, there appear to be major differences
    among the proportion of sites in which the scientists detected a track
    (Figure x). The difference in estimated na\"ive occupancy among the
    observers indicates that the scientists did not always correctly identify
    the tracks.

@@

    \subsection{False positive rates}
    The lack of consistency among the flights is strong evidence that the data
    contain false positives where scientists think they see a track when it does
    not exist. Not accounting for such a process can lead to major biases; as
    the number of flights increases, the occupancy rate estimate will approach
    one (Royle and Link 2006). We introduced a false detection process and used
    the observer as a covariate (by means of indicator variables) for both the
    probability of correctly identifying a track that exists ($p$) and the
    probability of identifying a track that does not exist ($E$). Including
    false positives introduces a duality whereby $(p=x,E=y,\psi=z)$ is as good a
    solution as $(p=y,E=x,\psi=1-z)$. To address this problem we assumed
    $p\gg0.5$ and $E\ll0.5$, an approach recommended by Royle and Link (2006).

@@

    We can use Bayesian statistics in the Hierarchical model as long as we set a
    prior for every parameter that we include. We can then use Markov Chain
    Monte Carlo (MCMC) to find a posterior distributions for every parameter. We
    run into a problem of duality with the probability of correct detection
    \(p\), and the probability of false detection \(E\). These two variables
    have bimodal posterior distributions, because \(p=x,E=y,\psi=z\) is the same
    as \(p=y,E=x,\psi=1-z\). Since we know that is reasonable for \(p\) to be
    greater than \(E\), we solved this problem by restricting the distribution
    of \(p\) to the range \(0.6\text{ to }1\), and \(E\) to \(0\text{ to }0.4\).

@@

    Figure x caption:
    The box plots show how the proportion of sites in which an observer saw
    tracks (``occupancy'') differs among the three observers. Only those dates
    in which all three observers flew are included.

>>>>

    \subsection{Descriptive statistics}
    In 2003, three different observers flew over the rivers. Assuming that no
    tracks are laid down between individuals' flights on the same day, all three
    people should have seen the same tracks. While there is some error in
    classifying which site the observer saw the actual track in, each observer
    should at least be seeing the same, or nearly the same, number of sites with
    tracks. However, we did not find that to be the case. If we look only at
    days that all three people observed, there appear to be major differences
    among the proportion of sites in which the scientists detected a track
    (Figure \ref{obsPlots}). The difference in estimated na\"ive occupancy among 
    the
    observers indicates that the scientists did not always correctly identify
    the tracks.

    \begin{figure}
        	\centering
	    \includegraphics[width=5in]{observerPlots.pdf}
	    \caption{The box plots show how the proportion of sites in which an 
	    observer saw tracks (``occupancy'') differs among the three observers in 
	    2003. Only those dates on which all three observers flew are included.}
	    \label{obsPlots}
    \end{figure}

@@

    \subsection{False positive rates}
    The lack of consistency among the flights is strong evidence that the data
    contain false positives where scientists think they see a track when it does
    not exist. Not accounting for such a process can lead to major biases; as
    the number of flights increases, the occupancy rate estimate will approach
    1 \cite{Royle2006}. Unlike most occupancy models where scientists assume 
    that detection of the species in question guarantees its presence \cite
    {MacKenzie2006}, we introduced a false detection process whereby an observer 
    can mark down the existence of a track when in reality there is none. This 
    false detection process can occur as long as there is no track, whether or 
    not the site is occupied.

@@

    We can use Bayesian statistics in the Hierarchical model as long as we set a
    prior for every parameter that we include. We can then use Markov Chain
    Monte Carlo (MCMC) to find a posterior distributions for every parameter. We
    run into a problem of duality with the probability of correct detection
    \(p\), and the probability of false detection \(E\). These two variables
    have bimodal posterior distributions, because \(p=x,E=y,\psi=z\) is the same
    as \(p=y,E=x,\psi=1-z\). Since we know that is reasonable for \(p\) to be
    greater than \(E\), we solved this problem by restricting the distribution
    of \(p\) to the range \(0.6\text{ to }1\), and \(E\) to \(0\text{ to }0.4\),  
    an approach recommended by Royle and Link \cite{Royle2006}.

@@
